---
title: "How does penguin body mass differ across species?"
author: "Carine Ishimwe"
date: "2026-09-24"
categories: [R, data-analysis, penguins]
execute:
  echo: true
---

## Question and data

How does body mass differ among Adelie, Chinstrap, and Gentoo penguins?
I will compare species summaries and visualize the distributions.

Data come from [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Kristen Gorman and Palmer Station Antarctica LTER.
The data are available under the CC0 licence and included in the R package.

Package citation: Horst, A. M., Hill, A. P., and Gorman, K. B. (2020).
*palmerpenguins: Palmer Archipelago (Antarctica) penguin data.*
[doi:10.5281/zenodo.3960218](https://doi.org/10.5281/zenodo.3960218).

## Load and inspect

The setup chunk suppresses package startup messages while keeping code visible.

```{r}
#| label: setup
#| message: false
library(palmerpenguins)
library(ggplot2)
```

Inspect the number of observations, the first rows, and missing body masses.

```{r}
#| label: inspect-data
penguins_data <- palmerpenguins::penguins
nrow(penguins_data)
head(penguins_data)
sum(is.na(penguins_data$body_mass_g))
```

## Prepare the data

Exclude observations missing body mass or species.
Keep the original dataset unchanged.

```{r}
#| label: clean-data
keep <- !is.na(penguins_data$body_mass_g) &
  !is.na(penguins_data$species)
penguins_clean <- penguins_data[keep, ]
nrow(penguins_clean)
nrow(penguins_data) - nrow(penguins_clean)
```

## Compare species

Calculate the mean body mass in grams for each species.
The counts show how many observations contribute to each group.

```{r}
#| label: summarize-mass
table(penguins_clean$species)
mass_summary <- aggregate(
  body_mass_g ~ species,
  data = penguins_clean,
  FUN = mean
)
knitr::kable(
  mass_summary,
  digits = 1,
  col.names = c("Species", "Mean body mass (g)")
)
```

## Visualize the distributions

A boxplot shows the median and spread, adding information beyond the means.

```{r}
#| label: fig-mass-by-species
#| fig-cap: "Body mass distributions by penguin species, excluding missing body masses."
ggplot(penguins_clean, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(x = "Species", y = "Body mass (g)") +
  theme_minimal()
```

## Interpretation

Gentoo penguins had the highest mean body mass at 5076.0 g,
compared with 3700.7 g for Adelie and 3733.1 g for Chinstrap.
Adelie and Chinstrap had similar means and strongly overlapping
body mass distributions.
Although Gentoo penguins were heavier on average, the overlapping
ranges show that not every Gentoo was heavier than every penguin
of the other species.

The analysis retained 342 penguins after excluding two observations
with missing body mass.
These results describe the sampled penguins and do not establish
a causal effect of species on body mass.

## AI assistance acknowledgment

I used OpenAI ChatGPT as a Socratic tutor to develop my understanding 
through questions, predictions, and feedback.